Simulating High-Dimensional Multivariate Data

### 2020-12-23 020-simulation-algorithm.Rmd
4th draft. Address Ed comments Part I.

We now describe `bigsimr::rvec`, our algorithm to generate random vectors, which mirrors the classical NORTA algorithm above with some modifications for rank-based dependency matching:

1. Pre-processing for nonparametric dependency matching.  
	(i) Convert ${\bf R_{Kendall}}$ or ${\bf R_{Spearman}}$ to the corresponding MVN correlation ${\bf R_{Pearson}}$.
	(ii) Check that ${\bf R_{Pearson}}$ is semi-positive definite.  
	(iii) If not, compute the nearest correlation matrix ${\bf \widetilde{R}_{Pearson}}$ to ${\bf R_{Pearson}}$ in the Frobenius norm.  
2. Gaussian copula construction.  
	(i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_{Pearson}})$.  
	(ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$.  
3. Quantile evaluations.  
	(i) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;

The pre-processing (Step 1) takes advantage of the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertKendall) or \@ref(eq:convertSpearman), respectively (implemented as `bigsimr::cor_covert`).

A complication often arises at this stage: the resultant matrix may become indefinite, either through numerical error or naturally occurring in the correlation conversion.
Researchers working in multivariate computation frequently encounter such difficulties and need to find a close positive (semi-)definite matrix.
A widely-available routine for this task in `R` is `matrix::nearPD`, though it is not suitable for high dimensions.
To overcome this issue, we introduce `bigsimr::nearPSD` --- an implementation of @QS2006's method that computes the nearest correlation matrix to a given symmetric matrix in the Frobenius norm.
This routine as many applications beyond our primary goal of random vector generation.

Once the target margins and algorithm inputs are determined, Steps 2 and 3 are essentially a NORTA algorithm with modern high-performance computing implementations.
Specifically, step 2i uses either the efficient multi-core multivariate normal simulator `mvnfast`[@Fasiolo2016] or Google's JAX python library for GPU acceleration of the Cholesky factorization and matrix multiplication.

## A note on NORTA in higher dimensions

@@

>>>>

## NORTA in higher dimensions

@@

- Discuss how the HD NORTA defective problem and the scalability issues lead to our algorithm
We now describe `bigsimr::rvec`, our algorithm to generate random vectors, which mirrors the classical NORTA algorithm above with some modifications for rank-based dependency matching:

1. Mapping step  
   i. Convert the target Spearman correlation matrix ${\bf R_{Spearman}}$ to the corresponding MVN Pearson correlation ${\bf R_X}$.   
   or  
i$^\prime$. Convert the target Kendall $\tau$ matrix ${\bf R_{Kendall}}$ to the corresponding MVN Pearson correlation ${\bf R_X}$.
2. Checking admissibility and nearest correlation computation
    (i) Check that ${\bf R_{X}}$ is a correlation matrix, a PSD matrix with 1's along the diagonal.   
	(ii) If ${\bf R_{X}}$ is a correlation matrix, the input matrix is *admissible* in this scheme.  
	Otherwise,  
(ii$^\prime$) Replace ${\bf R_{X}}$ with the nearest correlation matrix ${\bf \widetilde{R}_{X}}$, in the Frobenius norm.
3. NORTA step
	(i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_{X}})$.  
	(ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$.  
	(iii) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;

The *Mapping step* takes advantage of the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertKendall) or \@ref(eq:convertSpearman), respectively (implemented as `bigsimr::cor_covert`).

Our algorithm allows the generation high-dimensional data with arbitrary marginal distributions with a broad class of admissible Spearman correlation matrices and Kendall tau matrices. The admissible classes consist of the matrices that map to a Pearson correlation matrix. In particular, if we let X be MVN with N components and set

$S_P = \{ R_p : R_p is a correlation matrix for X \}$
$S_K = \{R_K : R_K is a Kendall tau matrix for a MVN \}$
$S_S = \{R_s : R_s is a Spearman correlation matrix for a MVN\}$

then there are 1-1 mappings between these sets.

So if a user of your package says "I want a 100 dimensional RV with NB marginals where R_s is the target Spearman correlation matrix", you can easily check whether R_S is an element of the "admissible set" S_S. The same if they say they want the same marginals but a Kendall matrix R_K. As is, you check (at least based on paper) whether the Spearman satisfies 2-D bounds which is ok for 2-D problems. So you can provide higher dimensional information about whether something can be done using your algorithm. Furthermore, I would guess that these sets of admissible S_K and S_S are not that restrictive - In particular R_p approx R_s for a MVN so that would suggest the admissible set S_S is not very restrictive because R_p can be any psd matrix with 1's on diagonal.

The final step implements a high-performance NORTA algorithm.
Specifically, step 3i uses either the efficient multi-core multivariate normal simulator `mvnfast`[@Fasiolo2016] or Google's JAX python library for GPU acceleration of the Cholesky factorization and matrix multiplication and steps 3ii, 3iii are parallelized over CPU cores.

### 2020-12-23 060-conclusion-discussion.Rmd
4th draft. Address Ed comments Part I.

>>>>

Researchers working in multivariate computation frequently encounter such difficulties and need to find a close positive (semi-)definite matrix.
A widely-available routine for this task in `R` is `matrix::nearPD`, though it is not suitable for high dimensions.
To overcome this issue, we introduce `bigsimr::nearPSD` --- an implementation of @QS2006's method that computes the nearest correlation matrix to a given symmetric matrix in the Frobenius norm.
This routine as many applications beyond our primary goal of random vector generation.

### 2020-12-23 index.Rmd
4th draft. Address JP feedback
4th draft. Address Ed comments Part I.

--- 
title: "Simulating High-Dimensional Multivariate Data"
subtitle: "using the bigsimr R Package"
author:
  - Alfred G. Schissler
  - Edward J. Bedrick
  - Alexander D. Knudson
  - Tomasz J. Kozubowski
  - Tin Nguyen
  - Anna K. Panorska
  - Juli Petereit
  - Walter W. Piegorsch
  - Duc Tran
bibliography: ["bigsimr.bib", "packages.bib"]
abstract:  It is critical to realistically simulate data when conducting Monte Carlo studies and evaluating quantitative methods. But measurements are often correlated and high dimensional in this era of big data, however, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers may resort to simulation designs that posit independence or perform arbitrary data transformations. This greatly diminishes insights into the empirical operating characteristics of any proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness. This article introduces the `bigsimr` R package that provides a flexible, scaleable procedure to simulate high-dimensional random vectors with given marginal characteristics and dependency measures. We describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix. Finally, we demonstrate the power of `bigsimr` by applying these functions to our motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=1212$ patients and dimension $d =1026$.
---

@@

With all this mind, we present a scaleable, flexible multivariate simulation algorithm. 
The crux of the method lies in the construction of a Gaussian copula, in the spirit of the NORTA procedure.
Further, we introduce the `bigsimr` R package that provides high-performance software implementing our algorithm.
The algorithm design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed-form relationships between dependency measures for the multivariate normal (MVN) distribution.

>>>>

--- 
title: "Simulating High-Dimensional Multivariate Data"
subtitle: "using the bigsimr R Package"
author:
  - Alfred G. Schissler
  - Edward J. Bedrick
  - Alexander D. Knudson
  - Tomasz J. Kozubowski
  - Tin Nguyen
  - Anna K. Panorska
  - Juli Petereit
  - Walter W. Piegorsch
  - Duc Tran
bibliography: ["bigsimr.bib", "packages.bib"]
abstract:  It is critical to realistically simulate data when conducting Monte Carlo studies and evaluating quantitative methods. But measurements are often correlated and high dimensional in this era of big data, however, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers may resort to simulation designs that posit independence or perform arbitrary data transformations. This article introduces the `bigsimr` R package that provides a flexible, scaleable procedure to simulate high-dimensional random vectors with arbitrary marginal distributions and a broad class of admissible Spearman correlation and Kendall $\tau$ matrices. We describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix. Finally, we demonstrate the power of `bigsimr` by applying these functions to our motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=1212$ patients and dimension $d =1026$.
---

@@

With all this in mind, we present a scaleable, flexible multivariate simulation algorithm. 
The crux of the method lies in the construction of a Gaussian copula, in the spirit of the NORTA procedure.
Further, we introduce the `bigsimr` R package that provides high-performance software implementing our algorithm.
The algorithm design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed-form relationships between dependency measures for the multivariate normal (MVN) distribution.
